---
title: "Data Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Helpful Packages

Most applications for the QOR method likely require data cleaning and manipulation of both typical data frames and spatial objects. The list of R packages below are likely to be helpful in these tasks, and are dependencies of the QOR package itself:

```{r packages}
library(tidyverse)
library(dplyr)
library(haven)
library(sf)
library(QOR)

# You would need your local path below to run the setup examples:
local_path <- "~/GitHub/Academic/QOR/" # example path
#local_path <- "/path/to/your/QOR/directory/"  # Change this to your local path and delete the above line
```

# Downloading Example Files

In the provided example, we download zipped (.zip file) versions of: 

- zip_code: National zip code shapefiles from the U.S. Census Bureau
- state_shape: North Carolina state boundary shapefile from the North Carolina Geospatial Data Clearinghouse
- district_shapes: North Carolina school district shapefiles from the National Center for Education Statistics (NCES)

We unzip these files for direct use in this tutorial. We also set the set a generous download timeout given that several files are quite large.

Users should note that a small, anonymized random sample of North Carolina voter registration data is included in the QOR package for demonstration purposes. This sample data is used in the address cleaning example below. For real applications, however, users will need to obtain their own voter registration data from a relevant governmental unit (or a paid provider).

```{r download}
# Increase timeout for large files
options(timeout = 1000)

# Define URLs (example: North Carolina 2022 data)
extracts <- "https://www.dropbox.com/scl/fi/epopkf7wqtgs9ymod0w3h/Example-Extracts.zip?rlkey=mfuzylz2iinljy3f5te86fc1p&st=26tify0g&dl=1"
zip_codes <- "https://www.dropbox.com/scl/fi/sq0mudsvm1g7c9ljbiye1/ZIP_2022.zip?rlkey=v9gqxsx0b1dli7lnoi95qwlyv&st=p7lprz1m&dl=1"
state_shape <- "https://www.dropbox.com/scl/fi/a8bn3ht1d67xd412zixf5/North_Carolina_State_Boundary.zip?rlkey=2j0at3vvts5i62trqvm4itpa5&st=0soqpq4j&dl=1"
district_shapes <- "https://www.dropbox.com/scl/fi/eh557z55pg9051dq4mves/SCHOOL_SY2022.zip?rlkey=ds8a7pocgj7evgsmvhsn9dyow&st=tecyt8yc&dl=1"

# Download shapefiles
download.file(url = zip_codes, destfile = paste0(local_path, "data-raw/Downloads/zip_codes_2022.zip"), mode = "wb", method = "auto")
unzip(zipfile = paste0(local_path, "data-raw/Downloads/zip_codes_2022.zip"), exdir = paste0(local_path, "data-raw/Extracted"))
download.file(url = state_shape, destfile = paste0(local_path, "data-raw/Downloads/state_shape.zip"), mode = "wb", method = "auto")
unzip(zipfile = paste0(local_path, "data-raw/Downloads/state_shape.zip"), exdir = paste0(local_path, "data-raw/Extracted"))
download.file(url = district_shapes, destfile = paste0(local_path, "data-raw/Downloads/district_shapes_2022.zip"), mode = "wb", method = "auto")
unzip(zipfile = paste0(local_path, "data-raw/Downloads/district_shapes_2022.zip"), exdir = paste0(local_path, "data-raw/Extracted")) 
```

# Cleaning Addresses

Before using QOR, it is imperative to ensure that your input data contain a minimal set of string address components:

- street
- city
- state

You may find that these components themselves must be built from other sub-components (e.g., street name, street number, apartment number, etc...). Below, we read-in our small random sample of anonymous voters before address processing:

```{r cleaning}
# Download anonymized voter registration extracts (Taken Jan. 1, 2022)
download.file(url = extracts, destfile = paste0(local_path, "data-raw/Downloads/Extracts_2022.zip"), mode = "wb", method = "auto")
unzip(zipfile = paste0(local_path, "data-raw/Downloads/Extracts_2022.zip"), exdir = paste0(local_path, "data-raw/Extracted"))
sample_2022 <- haven::read_dta(paste0(local_path, "data-raw/Extracted/Example Extracts/VR_Snapshot_2022_ALL_Anonymized.dta"))

# View a few raw observations
print(head(sample_2022, n = 10))
```

Then, we can prepare addresses for geocoding. Our example was written in Stata and saved as a .dta file, though address cleaning can occur in any programming language.

**Example workflow:**

- Read in raw data
- Rename address component variables to a standard format across the data years that you are using
- Convert all address components/subcomponents to strings (e.g., string variables for street, city, postalcode, and state)
- _Likely Needed:_ Construct the street address from subcomponents
- Remove leading/trailing whitespace in all variables
- Make sure to retain or generate a unique state voter (or other unit) id along with separate variables for street, city, postalcode, and state

```{r prepare_addresses}
## Read in cleaned sample
# We call an external Stata .do file (available on GitHub) for simplicity:
library(RStata)
options("RStata.StataPath" = "/usr/local/stata19/stata-se") # Path to your Stata executable. Use console version instead of GUI (stata-se instead of xstata-se)
suppressWarnings(stata(src = paste0(local_path, "data-raw/NCAddresses.do"), stata.version = 19.5))

# Note: The above line requires the RStata package (install.packages("RStata")) and a local installation of Stata
# If using GUI version (xstata-se), GTK warnings may appear but can be ignored if output is correct
```

# Next Steps

Different raw data structures will require unique approaches to produce a format compatible with QOR functions. However, With cleaned addresses, you are now ready to geocode your data and apply the QOR method. 
Your addresses should look similar to the following example input dataframe when correctly prepared for query():

```{r example_addresses, eval=TRUE, message=FALSE, warning=FALSE}
# Read the cleaned example addresses
example <- haven::read_dta(system.file("example_data", "sample_2022_addresses.dta", package = "QOR"))

# View a few cleaned addresses
print(head(example, n = 10))
```

Please see the [Getting Started](https://adam-p-shepardson.github.io/QOR/articles/getting-started.html) vignette and the function documentation for further guidance.

```{r hidden_setup_end, include=FALSE, eval=TRUE}
# Hidden code chunk for deleting downloaded files after use
local_path <- "~/GitHub/Academic/QOR/"

# Zipped files
files_to_delete <- c(paste0(local_path, "data-raw/Downloads/Extracts_2022.zip"), paste0(local_path, "data-raw/Downloads/zip_codes_2022.zip"), 
paste0(local_path, "data-raw/Downloads/state_shape.zip"), paste0(local_path, "data-raw/Downloads/district_shapes_2022.zip"))
file.remove(files_to_delete)

# Extracted files
names <- c("Example Extracts", "ZIP_2022", "North_Carolina_State_Boundary", "SCHOOL_SY2022")
for (name in names) {
  to_delete <- list.files(paste0(local_path, "data-raw/Extracted/", name), full.names = TRUE)
  file.remove(to_delete)
}
```