---
title: "State FIPS Codes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State FIPS Codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# State FIPS Code Reference

FIPS (Federal Information Processing Standards) codes are necessary for the QOR method to work when using NCES shapefiles. The `overlay()` and `recover()` functions require the `state_FIPS` parameter to correctly filter school districts.

The table below provides a complete mapping of 50 U.S. state two-letter abbreviations to their FIPS codes per the [Bureau of Labor Statistics](https://www.bls.gov/respondents/mwr/electronic-data-interchange/appendix-d-usps-state-abbreviations-and-fips-codes.htm):

```{r mappings, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# Creating state FIPS reference table
state_fips <- tibble::tibble(
  FIPS = c("01","02","04","05","06","08","09","10","12","13","15","16","17","18","19","20","21","22","23","24","25","26","27","28","29","30","31","32","33","34","35","36","37","38","39","40","41","42","44","45","46","47","48","49","50","51","53","54","55","56"),
  Abbreviation = c("AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA","HI","ID","IL","IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","WA","WV","WI","WY"),
  State = c("Alabama","Alaska","Arizona","Arkansas","California","Colorado","Connecticut","Delaware",
    "Florida","Georgia","Hawaii","Idaho","Illinois","Indiana","Iowa","Kansas","Kentucky",
    "Louisiana","Maine","Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri",
    "Montana","Nebraska","Nevada","New Hampshire","New Jersey","New Mexico","New York",
    "North Carolina","North Dakota","Ohio","Oklahoma","Oregon","Pennsylvania","Rhode Island",
    "South Carolina","South Dakota","Tennessee","Texas","Utah","Vermont","Virginia","Washington",
    "West Virginia","Wisconsin","Wyoming")
    )

# Display table
knitr::kable(state_fips, align = c("c", "c", "l"))
```
